hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper.Rmd
edits to discussion

**hatchR** is designed primarily as a tool for predicting fish phenology. To maintain focus on this core function, we provide minimal build-in data validation and vizualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensize validation checks. However, we include basic data-checking functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`). Intuitive functions are provided for users to apply models (either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function). Users can then apply these models to water temperature data (e.g. from a HOBO temp logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external vizualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows–particulalry for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.

@@

With **hatchR** we present a software ecosystem that bridges the analytical gap of predicting developmental phenology for wild fishes and develops a formal framework for applying effective value models from user-provided parameterizations. To do so, the software is bundled in two forms, a fully customizable R package, especially useful for running repetitive or complex analyses and a graphical-user-interface designed for ease of use and tasks that may only run once or a handful of times. Both applications allows users to import their data, run basic data checks, and apply historic model parameterizations for Salmonids or create their own species- or population specific parameterizations. Additionally, we provide substantial documentation on the **hatchR** website (<https://bmait101.github.io/hatchR/>) to walk users through basic to advanced applications of the two platforms.

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salinity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions. Another factor that could bias estimates from these models is that temperatures used from sensors do not reflect the geomorphic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.

To date, the application of the effective value framework has largely focused using species-specific models to predict phenology in wild environments (see @adelfio2024, @austin2019). However, we emphasize that the statistical relationships at the basis of these models are fundamentally reaction norms, meaning that both the intercept and slope of the responses of fishes to temperature are reflective of family- or population-level genetic variation, genetic x environment responses, or may be indicative of phylogenetic differences among species (@west-eberhard2003). For instance, while @sparks2017 did not find differences in the rates of development between the focal populations of their study, they did observe family-level genetic x environment interactions across different thermal regimes. Similarly, when they reparameterized their models using fish from western Alaska, they found differences in both the slope and intercept of the model relative to the original parameterization from @beacham1990, notably derived from multiple Canadian stocks. The western Alaskan fish generally developed more slowly than their southerly counterparts, consistent with cogradient variation (@conover2009, @sparks2022). In this sense, we encourage users not to only think about the end points of these models (days to development) but also how the statistical relationships they are premised on inform our understanding of micro- and and macro-evolutionary processes.

**BRYAN check the below**

The models presented in **hatchR** can be further customized in multiple ways beyond the use-cases provided above. For instance, while our models are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initiation of external feeding like eye-up or other embryological developmental stages (@velsen1980) or ***Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal***. Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided (as long as it integrates daily temperature), they could pass a different expression into `predict_phenology()` for additional customization. Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in accordance with the power law and these developmental models could be extended to other organisms beyond fishes.

@@

**hatchR** bridges the analytical gap for the prediction of developmental phenology for fishes in wild environments. It provides basic data checks and summarization, options for existing or custom model parameterization, and simple to complex phenology prediction frameworks. Importantly, it extends the effective value framework developed by @sparks2019 to a generalizable tool that can be applied to any fish species or population as long as the appropriate source data are available, which can easily be collected in aquaculture settings. We present basic applications for how **hatchR** may be used, with additional and more detailed examples provided on the software's [website](https://bmait101.github.io/hatchR/). **hatchR** is designed for applied and basic applications in which users can engage with the resource in a programmatic environment or a point-and-click Shiny app. We expect we have only demonstrated a small suite of the applications **hatchR** may be used for and encourage the user community to creatively build on the framework developed here.

>>>>

**hatchR** is designed primarily as a tool for predicting fish phenology. To maintain focus on this core function, we provide minimal build-in data validation and vizualization tools, as users are expected to understand and check their own data. Given the diversity of potential data types, it is impractical to implement comprehensize validation checks. However, we include basic data-checking functions (`check_continuous()`, `summarize_temp()`) and limited built-in visualization capabilities (`plot_check_temp()`, `plot_phenology()`). Intuitive functions are provided for users to apply models (either existing models from the literature using the `model_select()` function or fitting custom functions from data using the `fit_model()` function). Users can then apply these models to water temperature data (e.g. from a HOBO temp logger) to predict when hatching phenology will occur. This is accomplished with the `predict_phenology()` function. The R package provides example workflows for customizing plots from model output, while the Shiny application includes a default output plot and an option to download results for external visualizations. For a high-level overview of **hatchR**'s applications, see Figure 1. Additional details on key functions and workflows–particularly for automating phenology predictions across multiple variables–are available in articles hosted on the software's webpage.

@@

With **hatchR**, we present a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes. It establishes a formal framework for applying effective value models from user-provided parameterizations. The software is available in two formats:

* A fully customizable R package, ideal for complex and repetitive analyses.
* A graphical user interface (GUI) for ease of use, designed for tasks that may only need to be run once or a few times.

Both versions allow users to import data, perform basic data checks, and apply either pre-existing salmonid model parameterizations or generate custom models specific to other species or populations. To support users at various levels of expertise, we provide extensive documentation on the hatchR website (https://bmait101.github.io/hatchR/), covering basic and advanced applications.

## Assumptions and considerations in applying effective value models

The application of hatchR and the effective value modeling framework relies on several key assumptions. First, environmental stressors may alter developmental timing. While effective value models predict developmental timing based on temperature, studies have shown that stressful environmental conditions—such as low dissolved oxygen, altered pH, high salinity, pathogen exposure, or mechanical disturbance—can induce premature hatching or emergence (@quinn2018; @cowan2024). Users should consider how such factors may influence their predictions.

Second, developmental timing occurs as a distribution, not a fixed point. While **hatchR** provides point estimates for developmental phenology, fish spawning and development within populations occur as distributions rather than single events (@mason1976). We encourage users to predict phenology using early, peak, and late thresholds (e.g., 5th, 50th, and 95th percentiles) or incorporate real or modeled distributions to capture variation.

Third, sensor-based temperature data may differ from actual egg temperatures. Water temperatures recorded by environmental sensors may not fully reflect thermal conditions in spawning microhabitats, where geomorphic factors influence temperature regimes (@geist2002). Users should consider how differences between measured and actual incubation temperatures may affect predictions.

## Evolutionary and population-level considerations

To date, effective value models have primarily been used to predict phenology in wild environments using species-specific parameterizations (@adelfio2024; @austin2019). However, these models fundamentally represent reaction norms, meaning that temperature-development relationships are influenced by genetic variation, gene-environment interactions, and phylogenetic differences (@west-eberhard2003).

For example, @sparks2017 found no significant differences in developmental rates between populations in their study but did observe family-level genetic × environment interactions under different thermal regimes. Similarly, when they reparameterized their models using western Alaskan Sockeye Salmon, they found slower developmental rates compared to populations from Canada (@beacham1990), consistent with cogradient variation (@conover2009; @sparks2022). These findings highlight the importance of considering not only developmental time predictions but also how the underlying statistical relationships inform micro- and macro-evolutionary processes in fishes.

## Future directions OR Expanding the utility of **hatchR**

The models within **hatchR** can be customized in multiple ways beyond the examples provided. While our current framework focuses on predicting hatch or emergence timing, it could be adapted to other key developmental milestones, such as early embryonic stages (e.g., eye-up in salmonids; @velsen1980), initiation of pelagic-larval dispersal in marine fish, where temperature regulates the transition from demersal to pelagic behavior, or current-mediated dispersal in riverine species, where larvae begin downstream drift at temperature-dependent thresholds.

Additionally, while `fit_model()` uses non-linear regression to estimate parameters, `predict_phenology()` only requires users to provide a model expression. This means that users can integrate alternative model structures, as long as they incorporate daily temperature, allowing further customization of predictions.

Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power-law relationship with temperature. Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions.

@@

**hatchR** provides a versatile and accessible tool for predicting developmental phenology in wild fish populations. It offers basic data checks and summarization tools, pre-existing and customizable model parameterization options, and scalable applications from simple site-level predictions to complex multi-site analyses.

Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings. We present foundational applications of **hatchR**, with additional use cases and implementation guides available on the software's website. The software is designed for both applied and basic research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app. We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.
